---
title: "Statistical Inference Course Project - Part 1"
author: "Emily Carr"
date: "January 29, 2017"
output: html_document
---
***
# Part 1: Simulation Exercise  
  
### Overview
In Part 1, I'm investigating the **exponential distribution** and comparing it with the **Central Limit Theorem**. I'll use a large number of simulations to compare sample vs. theoretical mean & variance and determine whether the distribution is approximately normal.
The theoretical mean and variance will be that of the **standard normal distribution**, because we expect, according to the Central Limit Theorem, that with a large sample size of a distribution of exponential averages, the distribution will look more like a standard normal rather than an exponential.
  
  
### Simulations
Below I'll perform 1000 simulations of a distribution of averages of 40 exponentials, using $\lambda = 2$:
```{r}
n <- 40
lambda <- 2
set.seed(1)
expdistavg <- NULL
for (i in 1 : 1000) expdistavg <- c(expdistavg, mean(rexp(n, lambda)))
```
  
  
  
### Sample Mean vs. Theoretical Mean
The theoretical mean of the sampling distribution is $\mu$, which is $1/\lambda$ given the exponential distrbution parameters:
```{r}
mean_t <- 1/lambda
mean_t
```
The sample mean for the simulated distribution of averages is:
```{r}
mean_s <- mean(expdistavg)
mean_s
```
Overlaying these values on the graph shows how close the sample mean is to the theoretical mean. Knowing the **Central Limit Theorem**, this is what we would expect from such a large sample size:
```{r}
hist(expdistavg, probability = TRUE, main = "Distribution of Exponential Averages")
abline(v = mean_s, col = 4, lwd = 2, lty = 2)
abline(v = mean_t, col = 4, lwd = 2, lty = 1)
legend(x = "topright", legend = c("Sample Mean", "Theoretical Mean"), col = c(4, 4), lwd = c(2, 2), lty = c(2, 1), cex = 0.6)
```
  
  
  
### Sample Variance vs. Theoretical Variance
The theoretical variance of the sampling distribution is $\sigma^2/n$, which is $1/(n*\lambda^2)$ given the exponential distrbution parameters:
```{r}
var_t <- 1/(n*lambda^2)
var_t
```
The sample variance for the simulated distribution of averages is:
```{r}
var_s <- var(expdistavg)
var_s
```
The variability of the simulated distribution of averages is very close to the theoretical variance of a standard normal. It's clear that if we increased $n$, per the **Central Limit Theorem**, the variance will get closer and closer to 0, which is the variance of a **standard normal distribution**.
  
  
  
### Distribution
Plotting the distrubution of averages against the normal density curve shows it is approximately normal:
```{r}
hist(expdistavg, probability = TRUE, main = "Distribution of Exponential Averages")
curve(dnorm(x, mean = mean_t, sd = sqrt(var_t)), col = 3, lwd = 2, add = TRUE)
legend(x = "topright", legend = c("Normal Density Curve"), col = c(3), lwd = c(2), lty = c(1), cex = 0.6)
```
